---
title: "BS: Simulating Warmth Term Ratings"
description: "How to explore the statistics of temperature language without leaving your towel — a prior predictive simulation."
author: "Michael Friendly"
date: "2026-06-13"
draft: true
categories: [beach science, simulation, statistics]
---

One of the great advantages of Beach Science is that you can do the statistics without leaving your towel. No ethics board, no recruitment drive, no clipboards. Just a simulation that captures the essential structure of the data you would collect if you were so inclined.


## The Core Idea

We want to model what happens when 200 people are each asked to assign a temperature in °C to 25 words — from *freezing* to *broiling*. The real study would take weeks to run. The simulation takes seconds, and lets us explore the statistical structure before committing to anything.

The key insight is that not all words are equally precise. *Freezing* is anchored — nearly everyone agrees it means something near 0°C. *Mild* is a vessel — people pour their own thermal history into it. The simulation models this explicitly.


## The Terms and Their Parameters

Each term is assigned two numbers:

- **`mean_temp`** — the consensus central temperature in °C, from ~0°C (*Freezing*) to ~46°C (*Broiling*).
- **`precision`** — a weight from 0 (maximally vague) to 1 (maximally anchored). *Freezing* gets 0.9; *Mild* and *Comfortable* get 0.35. This is the most debatable part of the setup — and deliberately so.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(ggridges)
library(scales)

set.seed(1234)

terms <- tribble(
  ~term,          ~mean_temp, ~precision,
  "Freezing",      0,          0.9,
  "Frigid",        2,          0.8,
  "Icy",           3,          0.85,
  "Bitter",        4,          0.75,
  "Raw",           7,          0.6,
  "Chilly",       10,          0.7,
  "Cool",         14,          0.6,
  "Crisp",        13,          0.55,
  "Fresh",        15,          0.4,
  "Mild",         18,          0.35,
  "Temperate",    19,          0.45,
  "Comfortable",  21,          0.35,
  "Pleasant",     22,          0.35,
  "Neutral",      20,          0.5,
  "Lukewarm",     30,          0.6,
  "Tepid",        28,          0.55,
  "Warm",         24,          0.4,
  "Balmy",        26,          0.45,
  "Toasty",       28,          0.55,
  "Sultry",       32,          0.5,
  "Hot",          35,          0.75,
  "Sweltering",   38,          0.7,
  "Scorching",    42,          0.85,
  "Blistering",   44,          0.8,
  "Broiling",     46,          0.85
)
```


## The Variance Model

The standard deviation for each term is:

$$\text{sd} = (1 - \text{precision}) \times (3 + \sqrt{\text{mean\_temp}} \times 0.8)$$

This combines mean-scaling (hotter terms are slightly more variable) with a precision override (anchored terms stay tight regardless of mean).


## The Error Distribution

Ratings are drawn from a $t(4)$ distribution — heavier tails than normal. The Finn who calls 15°C warm, the Senegalese respondent for whom *mild* starts at 28°C. The tails are where the interesting individual differences live.


## Individual Differences

Two sources of systematic shift:

- **Geographic origin** (northern / temperate / southern): ~3°C shift across all ratings
- **Acclimatisation** (0–14 days): each beach day adds ~0.15°C to warmth perception

```{r}
#| label: subjects
n_subjects <- 200

subjects <- tibble(
  id          = 1:n_subjects,
  origin      = sample(c("northern", "southern", "temperate"),
                       n_subjects, replace = TRUE, prob = c(0.3, 0.3, 0.4)),
  acclim_days = sample(0:14, n_subjects, replace = TRUE)
) |>
  mutate(
    origin_shift = case_when(
      origin == "northern" ~ rnorm(n(), -3, 1),
      origin == "southern" ~ rnorm(n(), 3, 1),
      TRUE                 ~ rnorm(n(), 0, 1)
    ),
    acclim_shift = acclim_days * 0.15
  )
```

With subjects and terms defined, `crossing()` gives us all 200 × 25 = 5,000 ratings, each built from the term's mean, the subject's shifts, and a $t(4)$ error scaled by the term's variance:

```{r}
#| label: simulate
sim_data <- crossing(subjects, terms) |>
  mutate(
    term_sd = (1 - precision) * (3 + sqrt(mean_temp) * 0.8),
    error   = rt(n(), df = 4) * term_sd,
    rating  = mean_temp + origin_shift + acclim_shift * 0.3 + error,
    rating  = pmax(-10, pmin(60, rating))
  )

term_summary <- sim_data |>
  group_by(term, mean_temp, precision) |>
  summarise(med    = median(rating),
            mn     = mean(rating),
            sd     = sd(rating),
            .groups = "drop")

sim_data <- sim_data |>
  mutate(term = fct_reorder(term, rating, .fun = median))
```


## Simulation as Exploration

There is another reason to simulate before collecting data that does not get mentioned enough: it lets you play with the assumptions.

What if the $t$ distribution has 2 degrees of freedom instead of 4 — fatter tails, more extreme outliers? Change one number and rerun. What if geographic origin shifts ratings by 6°C rather than 3°C? What if acclimatisation matters twice as much as we assumed? Each of these is a one-line edit followed by a replot.

This is the what-if machine. You can explore the space of plausible worlds before committing to data collection in any of them. And crucially, when real data eventually arrives, you already have an intuition for what the results should look like — which means surprises are actually informative rather than just confusing.

Statisticians call this a *prior predictive check*. Beach Scientists call it an afternoon well spent.


### Ridgeline plot

```{r}
#| label: fig-ridgeline
#| fig-cap: "Simulated temperature ratings for 25 warmth terms (n = 200). White dot = median."
#| fig-width: 9
#| fig-height: 12
ggplot(sim_data, aes(x = rating, y = term, fill = mean_temp)) +
  geom_density_ridges(scale = 1.8, rel_min_height = 0.01,
                      colour = "white", linewidth = 0.3, alpha = 0.85) +
  geom_point(data = term_summary, aes(x = med, y = term),
             colour = "white", size = 1.8, inherit.aes = FALSE) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#d6604d",
                       midpoint = 22, name = "Mean\ntemp (°C)") +
  scale_x_continuous(breaks = seq(-10, 60, 10),
                     labels = function(x) paste0(x, "°")) +
  labs(title = 'How warm is "warm"?',
       subtitle = "Simulated ratings for 25 warmth terms (n = 200)\nWhite dot = median",
       x = "Assigned temperature (°C)", y = NULL,
       caption = "Beach Science | t(4) errors, variance scaled by precision") +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(plot.title = element_text(face = "bold", size = 16))
```


### Precision vs. spread

```{r}
#| label: fig-precision
#| fig-cap: "Observed SD of ratings vs. mean temperature, coloured by assigned precision."
#| fig-width: 8
#| fig-height: 6
ggplot(term_summary, aes(x = mean_temp, y = sd)) +
  geom_point(aes(colour = precision), size = 3) +
  geom_smooth(method = "loess", se = FALSE, colour = "grey60",
              linewidth = 0.7, linetype = "dashed") +
  ggrepel::geom_text_repel(aes(label = term), size = 3, max.overlaps = 20) +
  scale_colour_gradient(low = "#d6604d", high = "#2166ac",
                        name = "Precision\n(1 = anchored)") +
  labs(title = "Term precision vs. observed spread",
       x = "Mean temperature (°C)", y = "SD of ratings (°C)") +
  theme_minimal(base_size = 12)
```


### Geographic origin

```{r}
#| label: fig-origin
#| fig-cap: "Density of ratings for five contested terms, split by geographic origin. Dashed lines = group medians."
#| fig-width: 6
#| fig-height: 10
origin_summary <- sim_data |>
  filter(term %in% c("Mild", "Warm", "Comfortable", "Chilly", "Hot")) |>
  group_by(term, origin) |>
  summarise(med = median(rating), .groups = "drop")

sim_data |>
  filter(term %in% c("Mild", "Warm", "Comfortable", "Chilly", "Hot")) |>
  ggplot(aes(x = rating, fill = origin)) +
  geom_density(alpha = 0.5, colour = NA) +
  geom_vline(data = origin_summary, aes(xintercept = med, colour = origin),
             linewidth = 0.8, linetype = "dashed") +
  facet_wrap(~term, ncol = 1, scales = "free_y") +
  scale_fill_brewer(palette = "Set2", name = "Origin") +
  scale_colour_brewer(palette = "Set2", guide = "none") +
  labs(title = "Geographic origin shifts warmth perception",
       x = "Assigned temperature (°C)", y = "Density") +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "italic"), legend.position = "bottom")
```


## What This Isn't

The simulation assumes the precision weights and mean temperatures we gave it. A real study would estimate these from data. Maybe *lukewarm* is more anchored than we think. Maybe *warm* is more contested than *mild*.

The simulation is a prior. The study would be the update. That is, more or less, what Bayesian statistics is. And you can think about all of this lying on a beach towel, which is the point.
