---
title: "Mind Your Chunks"
subtitle: "The Care and Feeding of R Code Chunks"
description: "Treating a book's .qmd source files as data: parsing ~670 R code chunks into a queryable data frame to audit code-folding, and other problems"
author: "Michael Friendly"
date: today
draft: true
execute:
  eval: false
  warning: false
  message: false
categories: [Vis-MLM, quarto, knitr, books]
format:
  html:
    toc: true
    toc-depth: 3
---

This is another post in the [Making of Vis-MLM](../../../posts/2026-04-making-vis-mlm/index.qmd) series,
alongside [Quarto Is Not LaTeX, However Hard It Tries](../quarto-is-not-latex/index.qmd) and
[Creating Book Indexes in Quarto](../../../posts/2026-04-quarto-indexes/index.qmd). This one is about a smaller,
more mechanical problem: once a book has hundreds of code chunks, how do you even find the ones that need attention?

OK, you've written an R-based book with extensive code examples---all analyses and figures
are reproducible from source. But now, for publication in both online and print form, you ask:
is it really necessary to show **all that code inline**? Or, you see a code block in the PDF that extends outside the shaded box.

A reviewer of my book [*Visualizing Multivariate Data and Models in R*](https://friendly.github.io/Vis-MLM-book/)
commented that there was too much code on display, and that the book could be shortened
by folding or hiding some of it. My RA, Gavin Klorfine, noticed that there were a bunch of code chunks that strayed into the margin.
The book contains 281 figures, most generated directly from R code, among 669 chunks throughout the book.

The question was:

> Can I make it easier to analyse the properties of all these chunks, in a way that could make editing easier to resolve these and other issues?

One solution to this class of problems, which I implemented by guiding Claude 4.6, involves simply treating your text `.qmd` files as "data",
scanning these with RegExs to find chunks and their properties, and writing this information to a `chunks` dataset that can be used to analyze
and fix such problems.

Some aspects of this work may be useful to others, so I describe the problems and solution steps below.

## Folding code in HTML and PDF

With Quarto, it is easy to fold code in HTML using the YAML chunk option `code-fold`. In your source `.qmd` file, it looks like this:

```{{r}}
#| label: fig-my-plot
#| code-fold: true
#| code-summary: "Show the code"
#| fig-cap: "A scatterplot"
ggplot(data, aes(x, y)) + geom_point()
```

This renders a collapsed "Show the code" button in HTML that the reader can expand at will.
Setting `code-fold: show` makes code visible but still collapsible; `code-fold: false` forces
it always open.

But `code-fold` is an HTML-only feature---it has no effect on PDF output, where the code
will still appear in full. For print, the relevant option is `echo`. You can suppress code
entirely in PDF while keeping it foldable in HTML with a small `knitr` trick: put
`echo=knitr::is_html_output()` directly on the chunk header line (not as a `#|` comment,
which won't work), so the header line above becomes
`` ```{r echo=knitr::is_html_output()} `` , with the same `#|` options following it as before.

Now the code is hidden in PDF and folded-but-available in HTML. The two options work
independently: `echo=knitr::is_html_output()` controls whether code reaches the output at
all; `code-fold` controls how it is presented once it does.

<!-- TODO: maybe add a side-by-side screenshot of HTML folded vs PDF suppressed? -->

## What else is worth tracking?

Once you start thinking systematically about code chunks in a large project, other questions
come up naturally:

- **Label style**: Quarto recommends `#| label: my-chunk` (YAML-style, inside the chunk body),
  but old-style knitr labels `` ```{r my-chunk} `` are still valid. A book that started before
  Quarto may have a mix. It would be nice to know which files still use the old style.
- **Figure inventory**: which chunks produce figures (`fig-cap` set, label starts with `fig-`),
  what size are they (`fig-width`, `fig-height`, `out-width`)? This information can be more generally
  useful in editing the book.
- **Unlabeled chunks**: chunks without labels can't be cross-referenced and are harder to
  audit. How many are there?
- **Dead code**: chunks with `eval: false` don't run at all. This was likely designed when writing, but is useful to know when reviewing.
- **Code volume**: how many lines of non-trivial code are in each chunk? Longer chunks are
  the higher-payoff targets for folding.

## Building a chunk database

The idea is simple: treat your `.qmd` source files as data and use R to parse them.
Each code chunk becomes one tidy row in a data frame, with columns for the attributes mentioned above.

In R, code chunks start with a line matching `` ```{r `` and end at the next line that is
exactly ` ``` `. Everything in between is either a `#|` YAML option line or code. That
structure is regular enough to parse with base-R string tools like `grep()` and
map that over the text using `lapply()`. No specialized packages needed.

```r
parse_file_chunks <- function(file) {
  lines  <- readLines(file, warn = FALSE)
  starts <- grep("^```\\{r", lines)          # chunk openings

  lapply(starts, function(s) {
    # find the closing ```
    e <- s + which(grepl("^```\\s*$", lines[(s+1):length(lines)]))[1]

    chunk_lines <- lines[(s+1):e]
    opts <- chunk_lines[grepl("^#\\|", chunk_lines)]   # YAML options
    code <- chunk_lines[!grepl("^#\\|", chunk_lines) &
                        !grepl("^```\\s*$", chunk_lines)]

    list(
      label     = get_opt(opts, "label"),     # #| label: value
      code_fold = get_opt(opts, "code-fold"),
      echo_html = grepl("echo=knitr::is_html_output()", lines[s], fixed=TRUE),
      n_code    = sum(nchar(trimws(code)) > 0)
    )
  })
}
```

For a full book project like *Vis-MLM*, with 19 source files and several active child files,
this produces a tidy data frame of 669 chunks with 17 fields:

```{r}
#| eval: true
load("data/chunks.RData")
dplyr::glimpse(chunks)
```

The fields include: `chapter`, `file`, `line`, `label`, `has_label`, `old_style`,
`is_fig`, `echo_html`, `echo_header`, `echo_opt`, `code_fold`, `code_summary`,
`eval`, `fig_cap`, `out_width`, `n_code_lines`, and a derived `fold_status`.

## The `fold_status` field

The key derived field is `fold_status`, which combines `code_fold` and `echo_html` into a
single summary:

| `fold_status` | Meaning |
|---|---|
| `"both"` | `echo_html` + `code-fold: true`---fully treated |
| `"fold-only"` | `code-fold: true` but no `echo_html`---HTML folds, code still in PDF |
| `"echo-only"` | `echo_html` but no `code-fold`---PDF hidden, but code visible in HTML |
| `"fold-show"` | `code-fold: show`---always shown, but collapsible |
| `"fold-false"` | `code-fold: false`---explicitly kept visible |
| `"no-eval"` | `eval: false`---chunk doesn't run |
| `"neither"` | No fold treatment at all |

For the *Vis-MLM* book, the breakdown across 281 figure chunks is:

```{r}
#| eval: true
table(chunks$fold_status[chunks$is_fig])
```

Reading that: `neither` (227) are potential targets for future folding; `fold-only` (19) are
folded in HTML but code still appears in PDF — incomplete; `both` (14) are fully treated
(Gavin's audit, this round); the rest (`fold-show`, `fold-false`, `no-eval`) are chunks where
the current treatment is intentional rather than an oversight.

The 19 `fold-only` figure chunks are a useful finding in themselves: they were folded in HTML
(probably added before the book targeted PDF), but code still appears in the printed version.
Those were candidates for a follow-up editing pass.

## Practical uses

With `chunks.RData` in hand, a few lines of R answer questions that would otherwise require
manually scanning thousands of lines of source:

**Which figure chunks still need folding treatment?** (first few, of 227)
```{r}
#| eval: true
head(chunks[chunks$is_fig & chunks$fold_status == "neither",
       c("chapter", "label", "n_code_lines")])
```

**How many old-style chunk labels remain?**
```{r}
#| eval: true
sum(chunks$old_style, na.rm = TRUE)
```

**Which chapters have the longest unfolded figure code?**
```{r}
#| eval: true
library(dplyr)
chunks |>
  filter(is_fig, fold_status == "neither") |>
  group_by(chapter) |>
  summarise(total_lines = sum(n_code_lines), n_figs = n()) |>
  arrange(desc(total_lines))
```

**Cross-checking a reviewer's audit**: given a list of chunk labels a reviewer flagged
for folding, join against `chunks` to confirm each has been treated and report its
`fold_status` automatically---no manual searching through source files required.

<!-- TODO: could link to the annotate-audit.R script or show a snippet -->

**Finding wide code in PDF**

For the most part in writing, I tried to be careful about formatting code for readability. But not always.
Rather than reviewing every such chunk manually, there was a fix within the Quarto rendering process.

Code lines that overflow the shaded box in a PDF book have a root cause in LaTeX geometry:
the available width is set by `\textwidth` (determined by the document class and trim size)
minus the padding of the code-block environment.
For this book (`krantz2`, `letterpaper`, 10 pt, Fira Mono in a `tcolorbox`) the safe limit
works out to roughly **65 characters** per line, well below R's default `options(width = 80)`.

Two fixes operate at different levels:

*LaTeX fix (automatic):* Loading `fvextra` (a fancyvrb extension) and setting
`breaklines=true` on the `Highlighting` environment makes any line that exceeds the
box width wrap automatically, with a small continuation symbol---no source edits required.
Plain fancyvrb does *not* support `breaklines`; `fvextra` must be loaded first
(after fancyvrb, which Pandoc loads in its generated preamble):

```latex
\usepackage{fvextra}
\RecustomVerbatimEnvironment{Highlighting}{Verbatim} %
    {commandchars=\\\{\},breaklines=true,breakanywhere=true}
```

`breaklines=true` wraps at spaces; `breakanywhere=true` additionally breaks mid-token
for long strings or identifiers with no internal spaces.

*R-side fix (targeted):* For source lines you want to reformat explicitly, the script
[`issues/wide-code.R`](https://github.com/friendly/Vis-MLM-book/blob/master/issues/wide-code.R)
extends the chunk-scanning approach to measure line widths.
It expands tabs with the same `tabstop=4` rule that knitr uses, then flags every
code line exceeding the limit---but only in chunks that are actually **visible in the
PDF** (skipping `echo=knitr::is_html_output()`, `echo=FALSE`, and `#| echo: false` chunks).
The output gives file, line number, chunk label, width, and a text snippet so each
case can be located and fixed directly.

## Summing up

For a large Quarto/knitr book project, treating code chunks as data pays off quickly.
The parsing logic is straightforward---chunks have a regular open/close structure and
`#|` options are easy to extract. The resulting data frame supports systematic auditing,
reviewer cross-checks, and bulk edits that would be tedious to do by hand.

The full script ([`chunks.R`](https://gist.github.com/friendly/765761673f9f74ada34ffa8051080ecd)) is available as a GitHub gist.
Adapt the `active_files` list and the `child_files` vector to your own project structure,
and you should be able to build a similar database in a few minutes.

<!-- TODO: check whether knitr or any other package already provides something like this -->
