---
title: "Distribution of Determinants of 3×3 Matrices from {1,2,...,9}"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: united
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
```

## Introduction

This document explores an interesting combinatorial and statistical problem: What is the distribution of determinants of all 3×3 matrices that can be formed from the numbers 1 through 9, each used exactly once?

We answer this question in two ways:

1. **Exact solution**: Calculate the determinant for all 9! = 362,880 permutations
2. **Permutation distribution**: Generate a large number of random permutations and approximate the distribution

## Setup

```{r libraries}
library(ggplot2)
library(dplyr)

# Function to compute determinant from a vector of 9 numbers
det_from_vec <- function(v) {
  M <- matrix(v, nrow = 3, ncol = 3, byrow = TRUE)
  det(M)
}
```

## Exact Solution: All 9! Permutations

Computing exact distribution using all 362,880 permutations of the digits 1-9.

```{r exact-solution}
# Generate all permutations using arrangements package
library(arrangements)

# If arrangements package is not available, use base R
if (!requireNamespace("arrangements", quietly = TRUE)) {
  cat("Using base R permutation generation...\n")
  
  # Use gtools if available
  if (requireNamespace("gtools", quietly = TRUE)) {
    all_perms <- gtools::permutations(9, 9, 1:9)
    determinants_exact <- apply(all_perms, 1, det_from_vec)
  } else {
    cat("For exact solution, install 'arrangements' or 'gtools' package\n")
    cat("install.packages('arrangements')\n")
    cat("Proceeding with permutation approach only...\n")
    determinants_exact <- NULL
  }
} else {
  # Use arrangements package (more efficient)
  all_perms <- arrangements::permutations(1:9, k = 9)
  determinants_exact <- apply(all_perms, 1, det_from_vec)
}
```

## Random Permutation Distribution

Generating random permutation distribution as a Monte Carlo approximation.

```{r random-permutations}
B <- 100000  # Number of random permutations
set.seed(42)  # For reproducibility

determinants_random <- replicate(B, {
  v <- sample(1:9, 9, replace = FALSE)
  det_from_vec(v)
})
```

## Summary Statistics

### Exact Solution

```{r summary-exact}
if (!is.null(determinants_exact)) {
  cat("EXACT SOLUTION (all 9! permutations):\n")
  cat(sprintf("  Total permutations: %d\n", length(determinants_exact)))
  cat(sprintf("  Range: [%.1f, %.1f]\n", 
              min(determinants_exact), max(determinants_exact)))
  cat(sprintf("  Mean: %.2f\n", mean(determinants_exact)))
  cat(sprintf("  Median: %.2f\n", median(determinants_exact)))
  cat(sprintf("  SD: %.2f\n", sd(determinants_exact)))
  cat(sprintf("  Unique values: %d\n", length(unique(determinants_exact))))
  
  # Check for zero determinants
  n_zero <- sum(determinants_exact == 0)
  cat(sprintf("  Zero determinants: %d (%.2f%%)\n", 
              n_zero, 100 * n_zero / length(determinants_exact)))
}
```

### Random Permutation Sample

```{r summary-random}
cat("RANDOM PERMUTATION SAMPLE (B =", B, "):\n")
cat(sprintf("  Range: [%.1f, %.1f]\n", 
            min(determinants_random), max(determinants_random)))
cat(sprintf("  Mean: %.2f\n", mean(determinants_random)))
cat(sprintf("  Median: %.2f\n", median(determinants_random)))
cat(sprintf("  SD: %.2f\n", sd(determinants_random)))
cat(sprintf("  Unique values: %d\n", length(unique(determinants_random)))
```

## Visualizations

### Comparison of Both Methods

```{r plot-comparison, fig.width=10, fig.height=6}
if (!is.null(determinants_exact)) {
  df_exact <- data.frame(
    determinant = determinants_exact,
    method = "Exact (all 9! permutations)"
  )
  
  df_random <- data.frame(
    determinant = determinants_random,
    method = paste0("Random sample (B = ", format(B, big.mark = ","), ")")
  )
  
  df_combined <- rbind(df_exact, df_random)
  
  # Plot: Comparison of both methods
  p1 <- ggplot(df_combined, aes(x = determinant, fill = method)) +
    geom_histogram(aes(y = after_stat(density)), 
                   bins = 100, alpha = 0.6, position = "identity") +
    geom_density(aes(color = method), linewidth = 1, fill = NA) +
    scale_fill_manual(values = c("#3498db", "#e74c3c")) +
    scale_color_manual(values = c("#2c3e50", "#c0392b")) +
    labs(
      title = "Distribution of Determinants of 3×3 Matrices from {1,2,...,9}",
      subtitle = "Each matrix uses digits 1-9 exactly once",
      x = "Determinant",
      y = "Density",
      fill = "Method",
      color = "Method"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(face = "bold", size = 14),
      panel.grid.minor = element_blank()
    )
  
  print(p1)
}
```

### Exact Distribution (Detailed View)

```{r plot-exact, fig.width=10, fig.height=6}
if (!is.null(determinants_exact)) {
  df_exact <- data.frame(determinant = determinants_exact)
  
  p2 <- ggplot(df_exact, aes(x = determinant)) +
    geom_histogram(aes(y = after_stat(density)), 
                   bins = 120, fill = "#3498db", alpha = 0.7) +
    geom_density(color = "#2c3e50", linewidth = 1.2) +
    labs(
      title = "Exact Distribution: All 362,880 Permutations",
      subtitle = "Determinants of 3×3 matrices using digits 1-9 once each",
      x = "Determinant",
      y = "Density"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      panel.grid.minor = element_blank()
    )
  
  print(p2)
}
```

### Random Permutation Sample Only

```{r plot-random, fig.width=10, fig.height=6}
if (is.null(determinants_exact)) {
  df_random <- data.frame(determinant = determinants_random)
  
  p <- ggplot(df_random, aes(x = determinant)) +
    geom_histogram(aes(y = after_stat(density)), 
                   bins = 100, fill = "#e74c3c", alpha = 0.7) +
    geom_density(color = "#c0392b", linewidth = 1.2) +
    labs(
      title = "Distribution of Determinants (Random Permutation Sample)",
      subtitle = paste0("B = ", format(B, big.mark = ","), 
                       " random 3×3 matrices using digits 1-9 once each"),
      x = "Determinant",
      y = "Density"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      panel.grid.minor = element_blank()
    )
  
  print(p)
}
```

## Additional Analysis: Distribution Properties

### Quartiles

```{r quartiles}
if (!is.null(determinants_exact)) {
  cat("Quartiles (exact):\n")
  print(quantile(determinants_exact, probs = c(0, 0.25, 0.5, 0.75, 1)))
}
```

### Most Common Values

```{r frequency-table}
if (!is.null(determinants_exact)) {
  cat("Top 10 most frequent determinant values:\n")
  freq_table <- sort(table(determinants_exact), decreasing = TRUE)
  print(head(freq_table, 10))
}
```

### Symmetry Check

```{r symmetry}
if (!is.null(determinants_exact)) {
  if (requireNamespace("moments", quietly = TRUE)) {
    cat("Symmetry check:\n")
    cat(sprintf("  Skewness: %.3f\n", 
                moments::skewness(determinants_exact)))
    cat("  (0 = symmetric, >0 = right-skewed, <0 = left-skewed)\n")
  }
}
```

## Conclusions

This analysis reveals the distribution of determinants when all possible 3×3 matrices are formed from the digits 1-9, each used exactly once. The Monte Carlo approximation with random permutations closely matches the exact distribution, validating the sampling approach for similar problems where exact enumeration might be computationally prohibitive.
